---
title: "Bayesian Data Analysis for Medical Data"
subtitle: Just Another Example of Hierarchical Modeling
author      : Paola Berchialla
job         : Department of Clinical and Biological Sciences, University of Torino
output:
  ioslides_presentation:
    css: assets/css/ioslides.css
    logo: assets/img/dscb.jpg
    mathjax: local
    self_contained: no
    widescreen: yes
  beamer_presentation: default
  slidy_presentation: default
---

<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
 </style>
 

## Do you remember?

$$
\begin{align}
y_{ik} &\sim \textrm N(\mu_{ik},\sigma^2)\\
\mu_{ik} &= \beta_0 + u_k + \beta_1x_{ik} \\
\beta_0&\sim \textrm N(0,100)\\
\beta_1&\sim \textrm N(0,100)\\
\sigma^2 &\sim \textrm{Uniform}(0,100)\\
u_k & \sim  \textrm N(0,\sigma_u^2)\\
\sigma_u^2 &\sim \textrm{Uniform}(0,100)\\
\end{align}
$$
 

```{r, eval=FALSE}
library(lme4)

mod_lmer <- lmer(log(totCort)  ~ Implant + days + Implant:days + (1|Ring),
data = cortbowl, REML = TRUE)
```

## Accounting for more structure

- Random factors can be nested


```{r, echo=FALSE}
load('R\\cortbowl.RData')
cortbowl$totCortlog <- log(cortbowl$totCort)
cortbowl$Ring <- as.integer(as.factor(cortbowl$Ring)) 
cortbowl$Age_z <- scale(cortbowl$Age)

```

```{r, results='asis', echo=FALSE}
library(knitr)
print(kable(head(cortbowl)))
```

## Model formulation

$$
\begin{align}
y_{ikj} &\sim \textrm N(\mu_{ikj},\sigma^2)\\
\mu_{i;k;j} &= \beta_0 + u_k + v_j + \beta_1x_{ik} \\
\beta_0&\sim \textrm N(0,100)\\
\beta_1&\sim \textrm N(0,100)\\
\sigma^2 &\sim \textrm{Uniform}(0,100)\\
u_k & \sim  \textrm N(0,\sigma_u^2)\\
v_j & \sim  \textrm N(0,\sigma_v^2)\\
\sigma_u^2 &\sim \textrm{Uniform}(0,100)\\
\sigma_v^2 &\sim \textrm{Uniform}(0,100)\\
\end{align}
$$


## The classical approach with lmer
```{r}
library(lme4)

mod_lmer <- lmer(log(totCort)  ~ Implant + days + Age_z+ Implant:days + (1|Brood/Ring),
data = cortbowl, REML = TRUE)
```

## ... and Bayesian approach with myglmer2stan


```{r, eval=FALSE}
source('R\\myglmer2stan.R')
my_stan_rs_3 <- myglmer2stan(totCortlog ~ Implant + days + Age_z + Implant:days +  (1|Brood/Ring),
                             data = cortbowl,
                             calcWAIC = T,
                             warmup = 100, 
                             iter = 500, 
                             chains = 2) 
```
<span style="color:red">
Error in [[<-(`*`tmp`*`, name, value = "(Intercept)") : no such index at level 1
</span>

## Write down the model

```
data{
int N;
real totCortlog[N];
real ImplantP[N];
real days20[N];
real daysbefore[N];
real Age_z[N];
int Brood[N];	// look at the new data! 
int Ring[N];
real ImplantP_X_days20[N];
real ImplantP_X_daysbefore[N];
int N_Brood; // look at the new data!
int N_Ring;
}
```
## Write down the model
```
parameters{
real Intercept;
real beta_ImplantP;
real beta_days20;
real beta_daysbefore;
real beta_Age_z;
real beta_ImplantP_X_days20;
real beta_ImplantP_X_daysbefore;
real<lower=0> sigma;
real vary_Ring[N_Ring];
real vary_Brood[N_Brood]; // Hey! look at the new parameter!
real<lower=0> sigma_Ring;
real<lower=0> sigma_Brood; // Hey! look at the new parameter!
}
```

## Write down the model {.smaller}
```
model{
real vary[N];
real glm[N];
// Priors
Intercept ~ normal( 0 , 100 );
beta_ImplantP ~ normal( 0 , 100 );
...
sigma_Ring ~ uniform( 0 , 100 );
sigma_Brood ~ uniform( 0 , 100 );    // Hey! look at the new parameter!
sigma ~ uniform( 0 , 100 );
// Varying effects
for ( j in 1:N_Ring ) vary_Ring[j] ~ normal( 0 , sigma_Ring );
for ( j in 1:N_Brood ) vary_Brood[j] ~ normal( 0 , sigma_Brood ); // Hey! look at the new parameter!
// Fixed effects
for ( i in 1:N ) {
vary[i] <- vary_Ring[Ring[i]] + vary_Brood[Brood[i]]; // Hey! look at the new parameter!
glm[i] <- vary[i] + Intercept
+ beta_ImplantP * ImplantP[i]
+ beta_days20 * days20[i] + beta_daysbefore * daysbefore[i]
+ beta_Age_z * Age_z[i]
+ beta_ImplantP_X_days20 * ImplantP_X_days20[i]
+ beta_ImplantP_X_daysbefore * ImplantP_X_daysbefore[i];
}
totCortlog ~ normal( glm , sigma );
}
```

## Write down the model
```
generated quantities{
real dev;
real vary[N];
real glm[N];
dev <- 0;
for ( i in 1:N ) {
vary[i] <- vary_Ring[Ring[i]];
glm[i] <- vary[i] + Intercept
+ beta_ImplantP * ImplantP[i]
+ beta_days20 * days20[i]
+ beta_daysbefore * daysbefore[i]
+ beta_Age_z * Age_z[i]
+ beta_ImplantP_X_days20 * ImplantP_X_days20[i]
+ beta_ImplantP_X_daysbefore * ImplantP_X_daysbefore[i];
dev <- dev + (-2) * normal_log( totCortlog[i] , glm[i] , sigma );
}
}
```

## Create DSO object and sample

(1) Create a list with all your data
```{r, eval=FALSE}
mod_mat <- as.data.frame(model.matrix(~ Implant + days + Implant:days, data=cortbowl))
data_list <- list(N = nrow(cortbowl),
                  totCortlog = cortbowl$totCortlog,
                  ImplantP = mod_mat$ImplantP,
                  days20 = mod_mat$days20,
                  daysbefore = mod_mat$daysbefore,
                  Age_z = cortbowl$Age_z,
                  Ring = cortbowl$Ring,
                  ImplantP_X_days20 = mod_mat$`ImplantP:days20`,
                  ImplantP_X_daysbefore = mod_mat$`ImplantP:daysbefore`,
                  N_Ring = 151,
                  N_Brood = 54,
                  Brood = as.numeric((cortbowl$Brood))
)
```

## Create DSO object and sample

(2) put the model in an object string
(3) Create DSO object
```{r, eval = FALSE}

stan_dso <- stan_model(model_code=model_string)

```

(4) sample

```{r, eval=FALSE}
stanFit <- sampling(object = stan_dso, 
                    data = data_list, 
                    chains = 1, iter = 9000, 
                    warmup = 1000, thin = 1)
```

## Print the output{.smaller}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
source('R\\stanmer2.R')
load('output\\stanFit.RData')
source('R\\myglmer2stan.R')
library(rstan)
library(glmer2stan)
```

```{r}
stanmer2(stanFit)
```

## Launch Shinystan

```{r}
library(shinystan)
shinystan(stanFit)
```


## Getting the slides

* The slides for this course were created with Rmarkdown: [http://rmarkdown.rstudio.com/](http://rmarkdown.rstudio.com/). 
* They are available from [https://github.com/berkeley3/BDA-IZSTO](https://github.com/berkeley3/BDA-IZSTO).
* To re-compile the slides:

    + Download the directory containing the lectures from Github
    + In R open the .Rmd file and set the working directory to the lecture directory
    + Click the *KnitHTML* button on Rstudio or run the following commands: 
  
```{r RmarkdownChunk, eval=FALSE}
library(rmarkdown) 
render("main.Rmd")
